Factors affecting inferences on natural mortality and associated environmental effects in state-space age-structured assessment models

### 2025-08-08 Ecov_study/mortality/manuscript/manuscript.tex
saving work on project_0 revisions

  pdfauthor={Timothy J. Miller1,2; Greg Britten3; Elizabeth N. Brooks2; Gavin Fay4; Alex Hansell2; Christopher M. Legault2; Brandon Muffley5; John Wiedenmann6},

@@

\title{Factors affecting inferences on natural mortality and associated environmental effects in state-space age-structured assessment models}
\author{Timothy J. Miller\textsuperscript{1,2} \and Greg Britten\textsuperscript{3} \and Elizabeth N. Brooks\textsuperscript{2} \and Gavin Fay\textsuperscript{4} \and Alex Hansell\textsuperscript{2} \and Christopher M. Legault\textsuperscript{2} \and Brandon Muffley\textsuperscript{5} \and John Wiedenmann\textsuperscript{6}}
\date{09 June, 2025}

@@

\(^1\)corresponding author: \href{mailto:timothy.j.miller@noaa.gov}{\nolinkurl{timothy.j.miller@noaa.gov}}\\
\(^2\)Northeast Fisheries Science Center, Woods Hole Laboratory, 166 Water Street, Woods Hole, MA 02543 USA\\
\(^3\)Biology Department, Woods Hole Oceanographic Institution, 266 Woods Hole Rd. Woods Hole, MA, USA\\
\(^4\)Department of Fisheries Oceanography, School for Marine Science and Technology, University of Massachusetts Dartmouth, 836 S Rodney French Boulevard, New Bedford, MA 02740, USA\\
\(^5\)Mid-Atlantic Fishery Management Council, 800 North State Street, Suite 201, Dover, DE 19901 USA\\
\(^6\)Department of Ecology, Evolution, and Natural Resources. Rutgers University\\

@@

State-space population models are now used widely for fisheries stock assessment in Europe, the United States, and Canada \citep{nielsenberg14, cadigan16, pedersenberg17, stockmiller21}. Because application of these methods are considered best practice and recommended for the next generation of stock assessment models \citep{hoyleetal22, punt23}, it is expected their use will only grow globally. An appeal of state-space models lies in their formulation treating latent population characteristics as statistical time series with periodic observations that also may have error due to sampling or other sources of measurement error and therefore separating these sources of biological and measurement variability. Through advances in computational capacity, we can use sophisticated numerical approaches to estimate model parameters as mixed effects \citep{thorsonminto15, kristensenetal16}.

State-space stock assessment models, with non-linear functions of latent processes and numerous observation types with different probability distribution assumptions represent one of most complex classes of state-space models. The literature on the effects of various factors on reliabilty of inferences from state-space assessment models is growing \citep{lietal24, milleretal_inreview1}. The importance of contrast in population size and fishing mortality and quality of data used to fit assessment models including the state-space variety is known \citep{magnussonhilborn07, milleretal_inreview1}. Furthermore, estimation of natural mortality, and even temporal variablity is possible in many scenarios \citep{leeetal11, cadigan16, millerhyun18, milleretal_inreview1}.

@@

A critical aspect of fisheries assessment models and their use in management is short-term projections that are used to determine catch advice. While understanding drivers of recruitment is important particularly for subsequent effects on reference points, recruitment in short-term projections typically has little impact on the exploitable biomass in the first few projection years. However, assumptions for natural mortality (\(M\)) have immediate and larger effects on projected biomass because they affect the abundances at older age classes at the end of the data time series that constitute spawning biomass and catch \citep{brodziaketal08, stocketal21}.

Because of the effects of \(M\) on both biological reference points and short term projections, better understanding sources of variation in \(M\) would provide more accurate estimation of abundance and productivity and therefore improved management. Temporal variation in \(M\) is less studied than recruitment, but its importance for explaining variability in observations has been demonstrated in state-space assessment models for Atlantic cod and yellowtail flounder \citep{cadigan16, stocketal21}. \citet{derisoetal08} also demonstrated the importance of several factors affecting Pacific herring natural mortality.

Assessment models could include temporal variation in many aspects of population dynamics or how observations are related to the population. For example the Woods Hole Assessment Model (WHAM) can include process errors treated as random effects for transition in cohorts over time, catchability for indices of abundance, selectivity of fishing fleets or indices, movement between regions, or in \(M\) \citep{stockmiller21, milleretal_inreview}. However, misspecified temporal population process errors could lead to biased population and stock status estimation, and, therefore, poor fisheries management decisions \citep{legaultpalmer16, szuwalskietal18}. Studies of the reliability of inferences we might make regarding the presence of temporal variability in \(M\) are limited. \citet{milleretal_inreview1} found AIC could accurately distinguish process errors in the changes in abundance for cohorts, but not for those specifically due to \(M\) except when uncertainty in population observations (indices, catch and age composition) was low and there was greater temporal variation in natural mortality. In their simulation studies looking a models with multiple sources of process error, \citet{lietal24} found including more sources of process error than existed in the operating model was a better model-building approach than excluding them a priori.

Here we conduct a simulation study with operating models (OMs) varying by degree of observation error uncertainty, sources of process error, fishing history, temporal variation in environmental covariates, and magnitude of the effect of the covariate on natural mortality. The simulated observations from these OMs are fitted with estimating models that make alternative assumptions for sources of process error, and whether median M and covariate effects are estimated. We evaluate the effects of these factors on convergence of fitted models, whether Akaike's information criterion (AIC) can correctly determine the correct source of process error and correct assumption about covariate effects on natural mortality, and the degree of bias in inferences for relevant parameters and outputs of the assessment model.

@@

All of our analyses used the Woods Hole Assessment Model (WHAM) to construct both OMs and EMs \citep{millerstock20, stockmiller21, milleretal_inreview}. The WHAM package has been used extensively to configure OMs and EMs for several other simulation studies \citep{stocketal21, legaultetal23, lietal24, brittenetal_inreview, lietal_inreview_a} and is used to assess many commercially important stocks in the Northeast U.S. \citep[e.g.,][]{nefsc22, nefsc22a, nefsc24}. We used \href{https://github.com/timjmiller/wham/tree/77bbd946e4881216a439933473d1c58b21c270c3}{version 1.0.6.9000, commit 77bbd94} for to generate all results.

We completed a simulation study with a 288 operating models. The factors defining the configuration of each operating model which are described in detail in subsequent sections include source of population process error (3 levels), index and catch observation uncertainty (2 levels), environmental covariate uncertainty (2 levels), latent environmental covariate process stochasticity (4 levels), and fishing history (2 levels). We simulated 100 data sets for each operating model that included simulations of process errors.

For each simulated data set we fit a set of 12 estimating models (EMs). The factors that distinguish the estimating models which are also described in detail below include source of population process error type (3 levels) whether (median) \(M\) was estimated or assumed known (2 levels), and whether the effect of the environmental covariate on \(M\) was estimated or not (2 levels).

The sources of population process error that were used in the OMs or assumed in the EMs were on recruitment only (R), recruitment and changes in cohort abundance over time (R+S), or recruitment and \(M\) (R+M). We did not use the log-normal bias-correction feature for process errors or observations described by \citet{stockmiller21} for operating and EMs \citep{lietal_inreview}. Simulations were all carried out on the University of Massachusetts Green High-Performance Computing Cluster. Code for completing the simulations and summarizing results can be found at \url{https://github.com/timjmiller/SSRTWG/ecov_study/mortality}.

@@

In the WHAM model, environmental covariates are assumed to be described as state-space processes with annual observations of the true latent covariate \citep{milleretal16, stockmiller21}. In our simulations, the latent covariate is assumed to be a stationary first order autoregressive (AR1) process
\[
X_y|X_{y-1} \sim \text{N}\left(\mu_E\left(1-\rho_E\right) + \rho_E X_{y-1}, \left(1-\rho_E^2\right)\sigma^2_E\right)
\]
with marginal mean \(\mu_E=0\) and variance \(\sigma^2_E\). The four configuration of the latent environmental covariate in the operating models assume the one of two values for the marginal standard deviation \(\sigma_E \in \{0.1, 0.5\}\) and for the autocorrelation parameter \(\rho_E \in \{0, 0.5\}\).

The observations of the latent environmental covariate are assumed to be unbiased and Gaussian
\[
x_y|X_y \sim \text{N}\left(X_y,\sigma^2_e\right)
\]
The standard deviation of the environmental observations in the operating models is one of two values \(\sigma_e \in \{0.1, 0.5\}\). Figure \ref{om_ecov_example} provides example simulations of the latent process and observations under the alternative configurations.

@@

Many of the characteristics of the population biology and structure including the age classes (10 age classes (ages 1 to 10+)), time span (40 years), maturity (Figure \ref{om_inputs_fig}, top left), growth (Figure \ref{om_inputs_fig}, top right), time of spawning (1/4 of the year), and recruitment (Figure \ref{om_inputs_fig}, bottom right) are identical to \citet{milleretal_inreview1}. The maturity at age is a logistic function of age with age at 50\% maturity (\(a_{50}\)) = 2.89 and slope = 0.88 and weight at age is derived from a von Beralanffy growth function where \(t_0 = 0\), \(L_\infty = 85\), and \(k = 0.3\), and a length-weight relationship
\[
W_a = \theta_1 L_a^{\theta_2}
\]
where \(\theta_1 = e^{-12.1}\) and \(\theta_2 = 3.2\).

The general model for \(M\) in year \(y\) is a log-linear function of both covariate effects \(X_y\) and process errors \(\varepsilon_{M,y}\) and a parameter \(\beta_{M}\) that defines median \(M\)
\[
\log M_y = \beta_M + \beta_{E} X_y + \varepsilon_{M,y}
\]
where the process errors are modeled as random effects that may, in general, be autocorrelated normal random variables
\[
\varepsilon_{M,y}|\varepsilon_{M,y-1} \sim \text{N}\left(\varepsilon_{M,y-1},(1-\rho_M^2)\sigma_M^2\right)
\]
\citep{stockmiller21}, but we assume \(\rho_M = 0\) in our R+M OMs. We assume the median \(M\) rate \(e^{\beta_M} = 0.2\) is constant across ages. For R and R+S OMs and EMs, \(\varepsilon_{M,y} = 0\). For all R+M OMs, we assume the same standard deviation \(\sigma_M = 0.3\) and is estimated in the R+M EMs. The covariate effect is one of 3 alternative values in the operating models, \(\beta_\text{E} \in \{0,0.25,0.5\}\). The parameters defining the simulated covariate time series, size of the covariate effect, and any \(M\) random effects result in a range of different levels of variation in annual values (Figure \ref{M_example}).

We assumed expected recruitment each year is from a Beverton-Holt stock-recruit relationship (SRR)
\[
R_{y} = \frac{a \text{SSB}_{y-1}}{1 + b \text{SSB}_{y-1}}.
\]
All biological inputs to calculations of spawning biomass per recruit (i.e., weight, maturity, and \(M\) at age) are constant in the R and R+S OMs without covariate effects on \(M\). Therefore, steepness and equilibrium unfished recruitment are also constant over the time period for those OMs \citep{millerbrooks21}. As in \citet{milleretal_inreview1}, our assumed biological inputs and selectivity (defined below) with constant \(M\) result in equilibrium fishing mortality that reduces spawning biomass per recruit to 40\% of the unfished level is \(F_{40\%} = 0.348\). With an assumed unfished recruitment of \(R_0 = e^{10}\), setting \(\Fmsy = F_{40\%}\) results in a steepness of 0.69 and \(a=0.60\) and \(b = 2.4 \times 10^{-5}\). For R+M OMs and all OMs with covariate effects on \(M\), steepness is not constant, but we used the same \(a\) and \(b\) parameters as other operating models which equates to a steepness and \(R_0\) at the median of the time series processes for \(M\) and the covariate.

We also used the same two fishing scenarios as \citet{milleretal_inreview1} for OMs. In the first scenario, the stock experiences overfishing at 2.5\Fmsy for the first 20 years followed by fishing at \Fmsy for the last 20 years (denoted \(2.5\Fmsy \rightarrow \Fmsy\)). In the second scenario, the stock is fished at \Fmsy for the entire time period (40 years).

We configured all R, R+S, and R+M OMs with uncorrelated random effects on recruitment with standard deviation on log(recruitment) \(\sigma_R = 0.5\). This same assumption was used by \citet{milleretal_inreview1} for R+M OMs and other OMs with fishery selectivity and index catchability process errors. For R+S OMs, cohort temporal transition process errors were uncorrelated with \(\sigma_{2+} = 0.3\)

@@

We define the generation of observations of total catch, aggregate indices, and corresponding age composition identical to \citet{milleretal_inreview1}. There is a single fleet operating year round for catch observations with logistic selectivity for the fleet with \(a_{50} = 5\) and slope = 1 (Figure \ref{om_inputs_fig}, bottom left). Observations are generated for all 40 years of the model. There are two index time series intended to represent fishery-independent surveys occurring in the spring (0.25 way through the year) and the fall (0.75 way through the year). Catchability of both surveys are assumed to be 0.1. We assumed catch and index age composition observations aer generated from a logistic-normal distribution where errors on the multivariate normal scale are independent. The standard deviation parameter is also constant across ages.

Standard deviation for log-aggregate catch was 0.1. There were two levels of observation error variance for indices and age composition for both indices and fleet catch. A low uncertainty specification assumed standard deviation of both series of log-aggregate index observations was 0.1 and the standard deviation of the logistic-normal for age composition observations was 0.3. In the high uncertainty specification the standard deviation for log-aggregate indices was 0.4 and that for the age composition observations was 1.5. For all estimating models, standard deviation for log-aggregate observations was assumed known whereas that for the logistic-normal age composition observations was estimated.

@@

For each data set simulated from an operating model 12 estimating models were fit. There were three factors defining the configuration of each estimating model: 1) whether \(\beta_M\) was estimated or assumed known, 2) whether an environmental effect \(\beta_E\) was estimated or not, and 3) whether the process errors were assumed on recruitment only (R), recruitment and survival (R+S), or recruitment and \(M\) (R+M).

The configuration of the process errors in the estimating models generally matched the corresponding options in the operating models. For example, uncorrelated R+S was assumed for both the estimating and operating model. However, R+M EMs did not assume M random effects were uncorrelated (\(\rho_M\) was estimated). The environmental covariate observations were included in all estimation models to ensure comparability of AIC. All fixed effects parameters for selectivity, catchability, fully-selected fishing mortality, mean recruitment, initial abundance at age, and variances for logistic-normal age composition distributions were estimated. Any process error variance parameters for recruitment, survival, and \(M\) were also estimated. The observation error variance of the environmental observations and aggregate catch and indices were all assumed known at the true values.

@@

We measured the frequency of convergence when fitting each EM to the simulated data sets for each OM. There are various ways to assess convergence of the fit \citep[e.g.,][]{carvalhoetal21}, but we defined successful convergence as the Hessian of the marginal log-likelihood being invertible and providing variance estimates for the fixed effects parameters as recommended by \citet{milleretal_inreview1}. We calculated 95\% confidence intervals for probability of convergence using the Clopper-Pearson exact method \citep{clopperpearson34, thulin14}.

@@

We measured the frequency of correct model selection using marginal AIC. For a given operating model the set of models that were considered all made the same assumptions on whether or not to estimate \(\beta_M\) or it is assumed at the true value. For model \(m\), the marginal AIC is a function of the marginal log-likelihood maximized with respect to the fixed effects in the model \(\boldsymbol{\theta}\) and the number of fixed effects \(n\left(\boldsymbol{\theta}\right)\) estimated,
\[
\text{AIC}_m = -2\left[{\text{argmax}}_{\boldsymbol{\theta}} \log L_m\left({\boldsymbol{\theta}}\right) - n\left({\boldsymbol{\theta}}\right)\right].
\]
All model fits that successfully completed the optimization were used in these results. We used all of these fits because some lack of convergence would be expected for the correct behavior of more complicated models that include process errors that did not exist in the operating model. For example R+M EMs fit to R OMs would be expected to estimate no variance in the \(M\) random effects and the estimated variance parameter going to zero would cause poor convergence.

@@

All results here use OM simulations with fits that satisfied the convergence criterion described above. We used this conditioning, to reflect how practitioners would proceed in analyses of model fits with real assessment data. That is, practitioners would ensure models converged such that Hessian-based standard errors were available for all model parameter estimates.

We focused on statistical behavior of estimators of the covariate effect on \(M\) (\(\widehat \beta_E\)), the estimator of the median \(M\) parameter (\(\widehat \beta_M\)), and estimators of terminal year natural mortality \(\widehat M\), spawning stock biomass \(\widehat{\text{SSB}}\), and fully-selected fishing mortality rate \(\widehat F\). In preliminary analyses we results of the estimators of all the annual values for \(M\), SSB and fishing mortality over the whole time series, but we found no appreciable differences in patterns across the various factors defining the OMs and EMs. Furthermore, results for terminal year fishing mortality were generally inversely related to those for spawning stock biomass, and are provided in the Supplementary Materials and not discussed further.

We calculated median errors (ME) of \(\widehat \beta_E\), \(\widehat \beta_M\)), and the Hessian-based standard error estimators of these parameters (\(\widehat{SE}\left(\widehat \beta_E\right)\) and \(\widehat{SE}\left(\widehat \beta_M\right)\)). We calculated the median relative errors (MRE) of terminal year \(\widehat M\), \(\widehat{\text{SSB}}\), and \(\widehat F\). We also calculated the root mean square error (RMSE) and estimated probability of coverage of constructed 95\% confidence intervals for \(\widehat \beta_E\) and \(\widehat \beta_M\) for EMs that estimated these parameters. We constructed the confidence intervals for probabilities of CI coverage using the same methods as those for probabilities of convergence.

>>>>

  pdfauthor={Timothy J. Miller1,2; Gregory L. Britten3; Elizabeth N. Brooks2; Gavin Fay4; Alexander C. Hansell2; Christopher M. Legault2; Brandon Muffley5; John Wiedenmann6},

@@

\title{Factors affecting inferences on natural mortality and associated environmental effects in state-space age-structured assessment models}
\author{Timothy J. Miller\textsuperscript{1,2} \and Gregory L. Britten\textsuperscript{3} \and Elizabeth N. Brooks\textsuperscript{2} \and Gavin Fay\textsuperscript{4} \and Alexander C. Hansell\textsuperscript{2} \and Christopher M. Legault\textsuperscript{2} \and Brandon Muffley\textsuperscript{5} \and John Wiedenmann\textsuperscript{6}}
\date{18 June, 2025}

@@

\(^1\)corresponding author: \href{mailto:timothy.j.miller@noaa.gov}{\nolinkurl{timothy.j.miller@noaa.gov}}\\
\(^2\)Northeast Fisheries Science Center, Woods Hole Laboratory, 166 Water Street, Woods Hole, MA 02543 USA\\
\(^3\)Biology Department, Woods Hole Oceanographic Institution, 266 Woods Hole Rd. Woods Hole, MA, USA\\
\(^4\)Department of Fisheries Oceanography, School for Marine Science and Technology, University of Massachusetts Dartmouth, 836 S Rodney French Boulevard, New Bedford, MA 02744, USA\\
\(^5\)Mid-Atlantic Fishery Management Council, 800 North State Street, Suite 201, Dover, DE 19901 USA\\
\(^6\)Department of Ecology, Evolution, and Natural Resources. Rutgers University\\

@@

State-space population models are now used widely for fisheries stock assessment in Europe, the United States, and Canada \citep{nielsenberg14, cadigan16, pedersenberg17, stockmiller21}. Because application of these methods are considered best practice and recommended for the next generation of stock assessment models \citep{hoyleetal22, punt23}, it is expected their use will only grow globally. An appeal of state-space models lies in their separation of sources of biological and measurement variability by treating latent population characteristics as statistical time series with periodic observations measured with error. Through advances in computational capacity, we can use sophisticated numerical approaches to estimate model parameters as mixed effects \citep{thorsonminto15, kristensenetal16}.

State-space stock assessment models, with non-linear functions of latent processes and numerous observation types with different probability distribution assumptions represent one of most complex classes of state-space models. The literature on the effects of various factors on reliability of inferences from state-space assessment models is growing \citep{lietal24, milleretal_inreview1}. The importance of contrast in population size and fishing mortality (\(F\)) and quality of data used to fit assessment models including the state-space variety is known \citep{magnussonhilborn07, milleretal_inreview1}. Furthermore, estimation of natural mortality (\(M\)), and even temporal variability in \(M\) is possible in many scenarios \citep{leeetal11, cadigan16, millerhyun18, milleretal_inreview1}.

@@

A critical aspect of fisheries assessment models and their use in management is short-term projections that are used to determine catch advice. While understanding drivers of recruitment is important particularly for subsequent effects on reference points, recruitment in short-term projections typically has little impact on the exploitable biomass in the first few projection years. However, assumptions for \(M\) have immediate and larger effects on projected biomass because they affect the abundances at older age classes at the end of the data time series that constitute spawning biomass and catch \citep{brodziaketal08, stocketal21}.

Because of the effects of \(M\) on both biological reference points and short term projections, better understanding sources of variation in \(M\) would provide more accurate estimation of abundance and productivity and therefore improved management. Temporal variation in \(M\) is less studied than recruitment, but its importance for explaining variability in observations has been demonstrated in state-space assessment models for Atlantic cod and yellowtail flounder \citep{cadigan16, stocketal21}. \citet{derisoetal08} also demonstrated the importance of several factors affecting \(M\) for Pacific herring.

Assessment models could include temporal variation in many aspects of population dynamics or how observations are related to the population. For example the Woods Hole Assessment Model (WHAM) can include process errors treated as random effects for transition in cohorts over time (hereafter referred to as apparent survival), catchability for indices of abundance, selectivity of fishing fleets or indices, movement between regions, or in \(M\) \citep{stockmiller21, milleretal_inreview}. However, misspecified temporal population process errors could lead to biased population and stock status estimation, and, therefore, poor fisheries management decisions \citep{legaultpalmer16, szuwalskietal18}. Studies of the reliability of inferences regarding the presence of temporal variability in \(M\) are limited. \citet{milleretal_inreview1} found AIC could accurately distinguish process errors in apparent survival, but not for those specifically due to \(M\) except when uncertainty in population observations (indices, catch and age composition) was low and there was greater temporal variation in \(M\). In their simulation studies looking at models with multiple sources of process error, \citet{lietal24} found including more sources of process error than existed in the operating model was a better model-building approach than excluding them a priori.

Here we conduct a simulation study with operating models (OMs) varying by degree of observation error uncertainty, sources of process error, fishing history, temporal variation in environmental covariates, and magnitude of the effect of the covariate on \(M\). The simulated observations from these OMs are fitted with estimating models that make alternative assumptions for sources of process error, and whether median \(M\) and covariate effects are estimated. We evaluate the effects of these factors on convergence of fitted models, whether Akaike's information criterion (AIC) can determine the correct source of process error and correct assumption about covariate effects on \(M\), and the accuracy for estimators of relevant parameters and stock size and harvest rates derived from the assessment model.

@@

Our analyses used the Woods Hole Assessment Model (WHAM) to construct both OMs and EMs \citep{millerstock20, stockmiller21, milleretal_inreview}. The WHAM package has been used extensively to configure OMs and EMs for several other simulation studies \citep{stocketal21, legaultetal23, lietal24, brittenetal_inreview, lietal_inreview_a} and is used to assess many commercially important stocks in the Northeast U.S. \citep[e.g.,][]{nefsc22, nefsc22a, nefsc24}. We used \href{https://github.com/timjmiller/wham/tree/77bbd946e4881216a439933473d1c58b21c270c3}{version 1.0.6.9000, commit 77bbd94} to generate all results.

We completed a simulation study with 288 operating models. The factors defining the configuration of each operating model, described in detail in subsequent sections, include source of population process error (3 levels), index and catch observation uncertainty (2 levels), environmental covariate uncertainty (2 levels), temporal variation in the latent environmental covariate (4 levels), and fishing history (2 levels). We simulated 100 data sets for each operating model that included simulations of process errors.

For each simulated data set we fit a set of 12 estimating models (EMs). The factors that distinguish the estimating models, also described in detail below, include source of population process error type (3 levels) whether (median) \(M\) was estimated or assumed known (2 levels), and whether the effect of the environmental covariate on \(M\) was estimated or not (2 levels).

The sources of population process error that were used in the OMs or assumed in the EMs were on recruitment only (R), recruitment and apparent survival (R+S), or recruitment and \(M\) (R+M). We did not use the log-normal bias-correction feature for process errors or observations described by \citet{stockmiller21} for OMs and EMs \citep{lietal_inreview}. Simulations were all carried out on the University of Massachusetts Green High-Performance Computing Cluster. Code for completing the simulations and summarizing results can be found at \url{https://github.com/timjmiller/SSRTWG/ecov_study/mortality}.

@@

In the WHAM model, environmental covariates are assumed to be described as state-space processes with annual observations of the true latent covariate \citep{milleretal16, stockmiller21}. In our simulations, the latent covariate is assumed to be a stationary first order autoregressive (AR1) process
\[
X_y|X_{y-1} \sim \text{N}\left(\mu_E\left(1-\rho_E\right) + \rho_E X_{y-1}, \left(1-\rho_E^2\right)\sigma^2_E\right)
\]
with marginal mean \(\mu_E=0\) and variance \(\sigma^2_E\). The four configurations of the latent environmental covariate in the operating models assume one of two values for the marginal standard deviation \(\sigma_E \in \{0.1, 0.5\}\) and for the autocorrelation parameter \(\rho_E \in \{0, 0.5\}\).

The observations of the latent environmental covariate are assumed to be unbiased and Gaussian
\[
x_y|X_y \sim \text{N}\left(X_y,\sigma^2_e\right)
\]
The standard deviation of the environmental observations in the operating models is one of two values \(\sigma_e \in \{0.1, 0.5\}\). Figure \ref{om_ecov_example} provides example simulations of the latent covariate and observations under the alternative configurations.

@@

Many of the characteristics of the population biology and structure including the age classes (10 age classes (ages 1 to 10+)), time span (40 years), maturity (Figure \ref{om_inputs_fig}, top left), growth (Figure \ref{om_inputs_fig}, top right), time of spawning (1/4 of the year), and recruitment (Figure \ref{om_inputs_fig}, bottom right) are identical to \citet{milleretal_inreview1}. The maturity at age is a logistic function with age at 50\% maturity (\(a_{50}\)) = 2.89 and slope = 0.88 and weight at age is derived from a von Beralanffy growth function where \(t_0 = 0\), \(L_\infty = 85\), and \(k = 0.3\), and a length-weight relationship
\[
W_a = \theta_1 L_a^{\theta_2}
\]
where \(\theta_1 = e^{-12.1}\) and \(\theta_2 = 3.2\).

The general model for \(M\) in year \(y\) is a log-linear function of both covariate effects \(X_y\) and process errors \(\varepsilon_{M,y}\) and a parameter \(\beta_{M}\) that defines median \(M\)
\[
\log M_y = \beta_M + \beta_{E} X_y + \varepsilon_{M,y}
\]
where the process errors are modeled as random effects that may, in general, be autocorrelated normal random variables
\[
\varepsilon_{M,y}|\varepsilon_{M,y-1} \sim \text{N}\left(\varepsilon_{M,y-1},(1-\rho_M^2)\sigma_M^2\right)
\]
\citep{stockmiller21}, but we assume \(\rho_M = 0\) in our R+M OMs. We assume the median \(M\) rate \(e^{\beta_M} = 0.2\) is constant across ages. For R and R+S OMs and EMs, \(\varepsilon_{M,y} = 0\). For all R+M OMs, we assume the same standard deviation \(\sigma_M = 0.3\), which is estimated in the R+M EMs. The covariate effect is one of 3 alternative values in the operating models, \(\beta_\text{E} \in \{0,0.25,0.5\}\). The parameters defining the simulated covariate time series, size of the covariate effect, and any \(M\) random effects result in a range of different levels of variation in annual values (Figure \ref{M_example}).

We assumed expected recruitment each year is from a Beverton-Holt stock-recruit relationship (SRR)
\[
R_{y} = \frac{a \text{SSB}_{y-1}}{1 + b \text{SSB}_{y-1}}.
\]
All biological inputs to calculations of spawning biomass per recruit (i.e., weight, maturity, and \(M\) at age) are constant in the R and R+S OMs without covariate effects on \(M\). Therefore, steepness and equilibrium unfished recruitment are also constant over the time period for those OMs \citep{millerbrooks21}. As in \citet{milleretal_inreview1}, our assumed biological inputs and selectivity (defined below) with constant \(M\) result in equilibrium \(F\) that reduces spawning biomass per recruit to 40\% of the unfished level is \(F_{40\%} = 0.348\). With an assumed unfished recruitment of \(R_0 = e^{10}\), setting \(\Fmsy = F_{40\%}\) results in a steepness of 0.69 and \(a=0.60\) and \(b = 2.4 \times 10^{-5}\). For R+M OMs and all OMs with covariate effects on \(M\), steepness is not constant, but we used the same \(a\) and \(b\) parameters as other operating models which equates to a steepness and \(R_0\) at the median of the time series models for \(M\) and the covariate.

We also used the same two fishing scenarios as \citet{milleretal_inreview1} for OMs. In the first scenario, the stock experiences overfishing at 2.5\Fmsy for the first 20 years followed by fishing at \Fmsy for the last 20 years (denoted \(2.5\Fmsy \rightarrow \Fmsy\)). In the second scenario, the stock is fished at \Fmsy for the entire time period (40 years). The magnitude of the overfishing assumptions is intended to reflect estimates of overfishing for Northeast US groundfish stocks from \citet{wiedenmannetal19}.

We configured all R, R+S, and R+M OMs with uncorrelated random effects on recruitment with standard deviation on log(recruitment) \(\sigma_R = 0.5\). This same assumption was used by \citet{milleretal_inreview1} for R+M OMs and other OMs with fishery selectivity and index catchability process errors. For R+S OMs, apparent survival process errors were uncorrelated with \(\sigma_{2+} = 0.3\).

@@

We define the generation of observations of aggregate (total combined across ages) catch and indices, and corresponding age composition identical to \citet{milleretal_inreview1}. There is a single fleet operating year round for catch observations with logistic selectivity for the fleet with \(a_{50} = 5\) and slope = 1 (Figure \ref{om_inputs_fig}, bottom left). Observations are generated for all 40 years of the model. There are two index time series intended to represent fishery-independent surveys occurring in the spring (0.25 way through the year) and the fall (0.75 way through the year). Catchability of both surveys are assumed to be 0.1. We assumed catch and index age composition observations are generated from a logistic-normal distribution where errors on the multivariate normal scale are independent. The standard deviation parameter is also constant across ages.

Standard deviation for log-aggregate catch was 0.1. There were two levels of observation error variance for aggregate indices and age composition for both indices and fleet catch. A low uncertainty specification assumed standard deviation of both series of log-aggregate index observations was 0.1 and the standard deviation of the logistic-normal for age composition observations was 0.3. In the high uncertainty specification the standard deviation for log-aggregate indices was 0.4 and that for the age composition observations was 1.5. For all estimating models, the standard deviation for log-aggregate observations was assumed known at the true value whereas that for the logistic-normal age composition observations was estimated.

@@

Estimating models were fit to each of 100 simulated data sets from each operating model. There were three factors defining the configuration of each estimating model: 1) whether \(\beta_M\) was estimated or assumed known, 2) whether an environmental effect \(\beta_E\) was estimated or not, and 3) whether the process errors were assumed on recruitment only (R), recruitment and survival (R+S), or recruitment and \(M\) (R+M).

The configuration of the process errors in the estimating models generally matched the corresponding options in the operating models. For example, uncorrelated R+S was assumed for both the estimating and operating model. However, R+M EMs did not assume \(M\) random effects were uncorrelated (\(\rho_M\) was estimated). The environmental covariate observations were included in all estimation models, whether effects on \(M\) were estimated or not, to ensure comparability of AIC. All fixed effects parameters for selectivity, catchability, fully-selected \(F\), mean recruitment, initial abundance at age, and variances for logistic-normal age composition distributions were estimated. Any process error variance parameters for recruitment, survival, and \(M\) were also estimated. The observation error variance of the environmental observations and aggregate catch and indices were all assumed known at the true values.

@@

We measured the frequency of convergence when fitting each EM to the simulated data sets for each OM. There are various ways to assess convergence of the fit \citep[e.g.,][]{carvalhoetal21}, but we defined successful convergence as the Hessian of the marginal log-likelihood being invertible and providing variance estimates for the fixed effects parameters as recommended by \citet{milleretal_inreview1}. We calculated 95\% confidence intervals (CIs) for probability of convergence using the Clopper-Pearson exact method \citep{clopperpearson34, thulin14}.

@@

We measured the frequency of correct model selection using marginal AIC. For a given operating model the set of models that were considered all made the same assumptions on whether or not to estimate \(\beta_M\) or it is assumed at the true value. For model \(m\), the marginal AIC is a function of the marginal log-likelihood maximized with respect to the fixed effects in the model \(\boldsymbol{\theta}\) and the number of fixed effects \(n\left(\boldsymbol{\theta}\right)\) estimated,
\[
\text{AIC}_m = -2\left[{\text{argmax}}_{\boldsymbol{\theta}} \log L_m\left({\boldsymbol{\theta}}\right) - n\left({\boldsymbol{\theta}}\right)\right].
\]
All model fits that successfully completed the optimization were used for this set of analyses. We used all of these fits because some lack of convergence would be expected for the correct behavior of more complicated models that include process errors that did not exist in the operating model. For example R+M EMs fit to R OMs would be expected to estimate no variance in the \(M\) random effects and the estimated variance parameter going to zero would cause poor convergence.

@@

All results here use OM simulations with fits that satisfied the convergence criterion described above. We used this conditioning to reflect how practitioners would proceed in analyses of model fits with real assessment data. That is, practitioners would ensure models converged such that Hessian-based standard errors were available for all model parameter estimates.

We focused on statistical behavior of estimators of the covariate effect on \(M\) (\(\widehat \beta_E\)), the estimator of the median \(M\) parameter (\(\widehat \beta_M\)), and estimators of terminal year \(M\) (\(\widehat M\)), spawning stock biomass (\(\widehat{\text{SSB}}\)), and fully-selected \(F\) (\(\widehat F\)). In preliminary analyses we examined results for the estimators of all the annual values for \(M\), SSB and \(F\) over the whole time series, but we found no appreciable differences in patterns across the various factors defining the OMs and EMs. Furthermore, results for terminal year \(F\) were generally inversely related to those for spawning stock biomass, and are provided in the Supplementary Materials and not discussed further.

We calculated median errors (ME) of \(\widehat \beta_E\) and \(\widehat \beta_M\), and the Hessian-based standard error estimators of these parameters (\(\widehat{SE}\left(\widehat \beta_E\right)\) and \(\widehat{SE}\left(\widehat \beta_M\right)\)). We calculated the median relative errors (MRE) of terminal year \(\widehat M\), \(\widehat{\text{SSB}}\), and \(\widehat F\). We also calculated the root mean square error (RMSE) and estimated probability of coverage of constructed 95\% CIs for \(\widehat \beta_E\) and \(\widehat \beta_M\) for EMs that estimated these parameters. We constructed the CIs for probabilities of CI coverage using the same methods as those for probabilities of convergence.

The true values for terminal year SSB and \(F\) vary among simulations. For the \(i\)th simulated data the relative error for terminal year value \(\theta_i\) provided from the fitted estimation model is
\[
\text{RE}_i\left(\theta\right) = \frac{\widehat \theta_i - \theta_i}{\theta_i}
\]
We calculated RMSE as
\[
\text{RMSE}\left(\widehat \theta\right) = \sqrt{\frac{1}{n}\sum^n\left(\widehat \theta_i - \theta_i\right)^2}
\]
For ME and MRE of estimators, we constructed 95\% CIs using the binomial distribution approach as in \citet{stockmiller21} and \citet{milleretal_inreview1}. Because of the inverse relationship of the CI and the null hypothesis significance test, when the CI contains zero we do not have evidence of bias from the simulation study.
